---
title: "MTH6139 Time Series" 
subtitle: "Coursework 1" 
author: "Sami Rashid" 
date: "Spring term 2025" 
output: 
  html_document:
    toc: true
    toc_float: true
    theme: spacelab 
    highlight: tango
editor_options: 
  markdown: 
    wrap: 72
---

```{r, echo=FALSE}
# This code will display the QMUL logo at the top right of the page
# Do not change this code
htmltools::img(src = knitr::image_uri("images/QMlogo.png"),
               alt = 'logo',
               style = 'position:absolute; top:0; right:0; padding:10px; width:20%;')
```

------------------------------------------------------------------------

# **Introduction**

Bitcoin is a **cryptocurrency (virtual money)** designed to act as a
form of money and used to make payments outside the control of a group
or entity. Therefore, it removes the need for third-party involvement
such as banks and is **not controlled by the government**.

Being the **first-ever cryptocurrency**, it is an **extremely volatile
financial asset** and has seen large fluctuations in price since its
establishment in **2008**.

## **Purpose of the Project**

The purpose of this project is to:

> ✅ **Analyse Bitcoin's historical price** and how its value has
> changed over time.\
> ✅ **Use Prophet forecasting** to predict the future prices of
> Bitcoin.\
> ✅ **Identify seasonality patterns.**\

------------------------------------------------------------------------

# **Prophet Model**

In this section we identify *trends* and *forecast* the price of Bitcoin
using **Prophets** forecasting tool.

The dataset used in this project was obtained from **Kaggle**. It
contains the *daily* closing price of bitcoin. You can access the
dataset here:\
[Bitcoin Historical Price Dataset
(Kaggle)](https://www.kaggle.com/datasets/herculespeters/bitcoin-prices-cleaned-dataset)

## **Load Bitcoin Dataset**

First, we load the dataset using [`read_csv`]{style="color:purple"}.

```{r message=FALSE, warning=FALSE}
#Load dataset
library(readr)
BTC_USD <- read_csv("data/BTC-USD.csv")

#Check dataset is loaded correctly
head(BTC_USD)  
```

---

## **Formatting the Dataset**

To use the **Prophet** Model we first must:

-   Convert the date column to **Prophets** date format
    ([`ds`]{style="color:purple"})
-   Set the price as the variable ([`y`]{style="color:purple"})

```{r message=FALSE, warning=FALSE}
#Convert dataset to Prophet format
library(zoo)
BitcoinData <- data.frame(
  ds = zoo::as.yearmon(BTC_USD$Date), 
  y = BTC_USD$Price)

BitcoinData$ds <- as.Date(BitcoinData$ds)
```

---

## **Visualising Bitcoin Price**

Before we forecast Bitcoins price, it would be a good idea to *plot* a
chart of the **historical** price of Bitcoin.

For a clearer chart we use [`ggplot`]{style="color:purple"}

```{r message=FALSE, warning=FALSE}
#Plot Bitcoin Price History
plot(BitcoinData$ds, BitcoinData$y, 
     type = "l",
     main = "Bitcoin Price History",
     xlab = "Year", ylab = "Price (USD)")

```

---

## **Implement the Prophet Model**

The **Prophet** model will first identify any **trends or seasonality**
from the dataset.

```{r message=FALSE, warning=FALSE}
#Check dataset for weekly and yearly seasonality
library(prophet)
BitcoinModel <- prophet(BitcoinData, 
                        yearly.seasonality = TRUE, 
                        weekly.seasonality = TRUE)

#Forecast next 365 days
FutureBitcoin <- make_future_dataframe(BitcoinModel, periods = 365, freq = "day")
BitcoinForecast <- predict(BitcoinModel, FutureBitcoin)
```

---

## **Forecasting Bitcoin Price**

We can now [`plot`]{style="color:purple"} the **forecasted** price of
Bitcoin!

```{r}
#Plot Bitcoin Forecasted model
plot(BitcoinModel, BitcoinForecast,
     main = "Bitcoin Price Forecast",
     xlab = "Year", ylab = "Price (USD)")
```

The [`plot`]{style="color:purple"} shows the **historical** Bitcoin Price (black points) with the **forecasted** Bitcoin price (blue line). It also includes levels of **uncertainty** with the blue shaded area. 

The forecast suggests a price will **increase** with a continuous **upward trend**

--------------------------------------------------------------------------------

# **Seasonality**

**Prophet** can also give us insights into **weekly** and **yearly** seasonality.

```{r message=FALSE, warning=FALSE}
#Plots of Bitcoins seasonality
prophet_plot_components(BitcoinModel, BitcoinForecast)
```

From the *weekly* seasonality we can see that **Monday** has the **highest increase** in price whereas **Tuesday** has the **highest decrease**. 

Futhermore, from the *yearly* seasonality we can see a large peak during **January**. We also see big price increases around **June and December**. On the other hand, **April and October** show the largest decreases in price for the year. 

--------------------------------------------------------------------------------

